---
title: "Time Series Analysis"
author: "Jan Dudzik"
date: "6/2/2020"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# ARIMA model

### 1. Data description
Provided data shows Price Index of goods and services in Poland between years 2000 and 2020. The basic prices value reflects to December of 1999, and is set as a 100, so basically, every record is a mean value of selected month's prices in goods and services divided by mean value of this sector's prices in December 1999, and multiplied by 100.

```{r}
library(readxl)
prices_data <- read_excel("Szereg_niesezon.xls")

prices= ts(data=prices_data$TOWARYUSLUGI, frequency = 12,             
             start=c(2000,1), end=c(2020,2)) 


```

```{r, echo=FALSE}
library(ggplot2)
library(ggfortify)

autoplot(prices, main = "Level of prices (December 1999 = 100)", ts.colour = 'red',xlab="Year", ylab="Prices level", lty=3)
autoplot(diff(prices), main = "First differences in prices level (December 1999 = 100)", ts.colour = 'blue',xlab="Year", ylab="Difference of price level", lty=3)

```

According to initial plots, however this particular time series seems to be non-stationary, the first differences could be. Therefore I will examine that by Dickey-Fuller test.

### 2. Integration level

```{r, echo = FALSE}
source("functions04.R")
testdf(prices, 4)

``` 

Analysis of original time series has proven, that according to Dickey-Fuller Test, it is not stationary (p-value > 10 percent), that proves visual conclusion. What is more, according to Breusch-Godfrey test with no lags, the null hypothesis about no autocorrelation of residuals is rejected.

```{r, echo = FALSE}
testdf(diff(prices), 4)

``` 

Statistical tests of first differences of selected time series proves, that according to Dickey-Fuller test, we cannot reject null hypothesis of non-stationarity (p-value < 1 percent), and by p-value = 0.7993760 of Breusch-Godfrey test, there is no autocorrelation between residuals in this model 


